---
title: "Course Project 2: Impact of Severe weather events in US"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Synopsis

The goal of this project is what kinds of weather events in United Stated results in most damagable effect. I will use U.S National Oceanic and Atmospheric Administration's (NOAA) storm database. Although the data includes from 1950, I will use the most recent one (from 1990) considering that the earlier year lack good records. The main quetions for this project is two:

- what weather events are most hamful to population health
- what weather events have the greatest economic consequence

For the first question, I will use the data of the number of *fatalities* and *injuries*. Also, for the second question, I will use the data of *PROPDM* and *CROPDM*. Here is the documentation related to the description of the original data ([link](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf))


## Data Processing
```{r cache=TRUE}
#library(R.utils)
if (!"data.csv.bz2" %in% dir("./")) {
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "data.csv.bz2")
#        bunzip2("data.csv.bz2", overwrite=T, remove=F)
}
#df<-read.csv("data.csv")
df<-read.csv("data.csv.bz2")
dim(df)
```


## Overall histogram
```{r overall_plot}
## extract year information
df$year<-as.numeric(format(as.Date(df$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
hist(df$year, xlab="Year",main="Weather event freauency")
```

More recent years should considered more complete because old data missed many data. Thus, we will only use the data from 1990-2011.

```{r}
## More recent years should be considered more complete. -->1990-2011
index<-df$year>=1990
final<-df[index,]
dim(final)
```


## Results

#### 1. Across the United States, which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health?
```{r}
library(plyr)
library(ggplot2)
fat<-ddply(final,.(EVTYPE),summarize,sum=sum(FATALITIES))
fat_order<-fat[order(fat$sum,decreasing = TRUE),]
fat_order<-head(fat_order[fat_order$sum!=0,],10)
fat_order$EVTYPE <- factor(fat_order$EVTYPE, levels = fat_order$EVTYPE[order(fat_order$sum,decreasing = TRUE)])
inj<-ddply(final,.(EVTYPE),summarize,sum=sum(INJURIES))
inj_order<-inj[order(inj$sum,decreasing = TRUE),]
inj_order<-head(inj_order[inj_order$sum!=0,],10)
inj_order$EVTYPE <- factor(inj_order$EVTYPE, levels = inj_order$EVTYPE[order(inj_order$sum,decreasing = TRUE)])
```


```{r plot}
g1 <- ggplot(fat_order,aes(EVTYPE,sum))+geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))+
        labs(x="Event Type")+
        labs(title="Fatalities")
g2 <- ggplot(inj_order,aes(EVTYPE,sum))+geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))+
        labs(x="Event Type")+
        labs(title="Injuries")
g1
g2
```

It seems that *Excessive Heat* and *Tornado* dominately affect the number of *fatalities* of people in US. At the same time, *Tornado* also affect the number of *injuriety* dominately.  


#### 2. Across the United States, which types of events have the greatest economic consequences?

```{r question2}
index1<-final$PROPDMGEXP=="H"
index2<-final$PROPDMGEXP=="K"
index3<-final$PROPDMGEXP=="M"
index4<-final$PROPDMGEXP=="B"

final[index1,]$PROPDMG<-final[index1,]$PROPDMG*10^2
final[index2,]$PROPDMG<-final[index2,]$PROPDMG*10^3
final[index3,]$PROPDMG<-final[index3,]$PROPDMG*10^6
final[index4,]$PROPDMG<-final[index4,]$PROPDMG*10^9

index1<-final$CROPDMGEXP=="H"
index2<-final$CROPDMGEXP=="K"
index3<-final$CROPDMGEXP=="M"
index4<-final$CROPDMGEXP=="B"

final[index1,]$CROPDMG<-final[index1,]$CROPDMG*10^2
final[index2,]$CROPDMG<-final[index2,]$CROPDMG*10^3
final[index3,]$CROPDMG<-final[index3,]$CROPDMG*10^6
final[index4,]$CROPDMG<-final[index4,]$CROPDMG*10^9

pro<-ddply(final,.(EVTYPE),summarize,sum=sum(PROPDMG))
pro_order<-pro[order(pro$sum,decreasing = TRUE),]
pro_order<-head(pro_order[pro_order$sum!=0,],10)
pro_order$EVTYPE <- factor(pro_order$EVTYPE, levels = pro_order$EVTYPE[order(pro_order$sum,decreasing = TRUE)])

cro<-ddply(final,.(EVTYPE),summarize,sum=sum(CROPDMG))
cro_order<-cro[order(cro$sum,decreasing = TRUE),]
cro_order<-head(cro_order[cro_order$sum!=0,],10)
cro_order$EVTYPE <- factor(cro_order$EVTYPE, levels = cro_order$EVTYPE[order(cro_order$sum,decreasing = TRUE)])
```

```{r plot2}
g3<-ggplot(pro_order,aes(EVTYPE,sum))+geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))+
        labs(x="Event Type")+
        labs(title="Property Damage")
g4<-ggplot(cro_order,aes(EVTYPE,sum))+geom_bar(stat="identity")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1))+
        labs(x="Event Type")+
        labs(title="Crop Damage")
g3
g4
```

The most damagable severe event is *Flood* while another most damagable event is *Drought*.

## Conclusions

In conclusion, I found that:

- *Excessive heat* and tornado are most harmful for facilities and *tornado* is most harmful for injuries.
- *Flood* is most harmful to property damage and *drought* is most harmful to crop damage in US.


